Adaptation of Palmer amaranth to cropping systems

### 2021-07-26 Frontiers/Frontiers.Rmd
as

Palmer amaranth (*Amaranthus palmeri* S. Watson) is currently considered one of the most economically damaged weed species to cropping systems in the United States [@ward2013]. The species has showed a remarkable capacity to evolve resistance to herbicides. Palmer amaranth has evolved resistance to eight herbicide sites of action, increasing the weed management complexity [@lindsay2017]. Uncontrolled Palmer amaranth in competition for water, light and nutrients can drastically reduce crop yields [@berger2015]. Palmer amaranth is documented with potential to reduce 91%, 68%, and X% of corn [@massinga2001], soybean [@klingaman1994], and cotton yields.

Palmer amaranth is a fast growing summer annual forb indigenous to Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with species biology. The increased use and movement of farm equipment across locations, conservation agriculture (e.g., no-till), and reliance on herbicides for weed management are the main human mediated selection of Palmer amaranth to cropping systems. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@ward2013]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging herbicide resistant alleles among plants [@oliveira2018]. Also, Palmer amaranth small seed size (1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@yu2021], animals [@farmer2017], and plant propagules [@hartzler2016]. Therefore, Palmer amaranth dispersal capacity make the species one of the most successful cases of weed adaption to cropping systems.

Light and temperature are likely the main environment requirements for Palmer amaranth successful adaptation. Palmer amaranth is reported with an extended germination period [@jha2010]. Germination of Palmer amaranth is triggered by 18 C soil temperature [@keeley1987], and optimal germination and biomass production were at 35/30 C day and night temperatures [@guo2003]. Water has not shown to limit Palmer amaranth fitness. Under continuous water stress, Palmer amaranth survived and produced at least 14000 seeds plant-1 [@chahal2018]. Also, seeds from Palmer amaranth growing under water stress conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. The continuous global temperature warming can impact agriculture and promote niches for Palmer amaranth invasion/adaptation into new environments. Currently, the greatest climatic risk of Palmer amaranth establishment is agronomic crops in Australia and Sub-Sahara Africa [@kistner2018]. Temperature is a key factor limiting Palmer amaranth northward range expansion [@briscoerunquist2019]; however, under future climate change Palmer amaranth is likely to expand northward into Canada and Northern Europe [@kistner2018].

Palmer amaranth is already found in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the US, Palmer amaranth is commonly found at crop [@garetson2019] and non-crop land [@bagavathiannan2016] in the warm southern United States but its range is expanding to cool temperatures northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a],  Michigan [@kohrt2017], and Connecticut [@aulakh2021]. Successful cases of Palmer amaranth invasion and near to eradication is well documented in Minnesota [@yu2021] and Iowa. No Palmer amaranth actively growing was found in Canada; however, Palmer amaranth seeds was detected in sweet potato slips [@page2021]. Nonetheless, it seems fated to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Therefore, strategies on Palmer amaranth management should encompass the agroecosystem level but not attempts to eradicate the weed. The continuous Palmer amaranth dispersal and potential establishment into northern US/Canada warrant investigation on species growing morphology in such environments.

Most Palmer amaranth studies are based on reactive (e,g, herbicide and tillage) rather than proactive management. Understanding Palmer amaranth biology and growing strategies under different environments can enhance our knowledge on species adaptation. Also, it can aid on designing proactive and ecological tactics to limit Palmer amaranth range expansion and reduce its negative impact while conserving diversity. Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under different environments and timings across five locations.

>>>>

Palmer amaranth (*Amaranthus palmeri* S. Watson) is currently considered one of the most economically damaged weed species to cropping systems in the United States [@ward2013]. The species has showed a remarkable capacity to evolve resistance to herbicides. Palmer amaranth has evolved resistance to eight herbicide sites of action, increasing the weed management complexity [@lindsay2017]. Uncontrolled Palmer amaranth in competition for water, light and nutrients can drastically reduce crop yields [@berger2015]. Palmer amaranth is documented with potential to reduce 91%, 68%, and 54% of corn [@massinga2001], soybean [@klingaman1994], and cotton [@morgan2001] yields.

Palmer amaranth is a fast growing summer annual forb indigenous to Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with species biology. Farm mechanization, conservation agriculture (e.g., no-till), and reliance on herbicides for weed management are the main human mediated selection of Palmer amaranth to cropping systems. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@ward2013]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging herbicide resistant alleles among plants [@oliveira2018]. Also, Palmer amaranth small seed size (1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@yu2021], animals [@farmer2017], and plant propagules [@hartzler2016]. Therefore, Palmer amaranth dispersal capacity make the species one of the most successful cases of weed adaption to cropping systems.

Light and temperature are likely the main environment requirements for Palmer amaranth successful adaptation. Palmer amaranth is reported with an extended germination period [@jha2010]. Germination of Palmer amaranth is triggered by 18 C soil temperature [@keeley1987], and optimal germination and biomass production occur at 35/30 C day and night temperatures [@guo2003]. Water has not shown to limit Palmer amaranth fitness. Under continuous water stress, Palmer amaranth survived and produced at least 14000 seeds plant-1 [@chahal2018]. Also, seeds from Palmer amaranth growing under water stress conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. The continuous global temperature warming can impact agriculture and promote niches for Palmer amaranth invasion/adaptation into new environments. Currently, it is estimated that the greatest climatic risk of Palmer amaranth establishment is agronomic crops in Australia and Sub-Sahara Africa [@kistner2018]. Temperature is a key factor limiting Palmer amaranth expansion to cooler geographies [@briscoerunquist2019]; however, under future climate change Palmer amaranth is likely to expand northward into Canada and Northern Europe [@kistner2018].

Palmer amaranth is already documented in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the US, Palmer amaranth is established at crop [@garetson2019] and non-crop land [@bagavathiannan2016] in the warm southern United States but its range is expanding to cool temperatures northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a],  Michigan [@kohrt2017], and Connecticut [@aulakh2021]. Successful cases of Palmer amaranth invasion and near to eradication is reported in Minnesota [@yu2021] and Iowa. No Palmer amaranth actively growing was found in Canada; however, Palmer amaranth seeds was detected in sweet potato slips [@page2021]. Nonetheless, it seems fated to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Therefore, strategies on Palmer amaranth management should encompass the agroecosystem level but not attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term ecological management.

The continuous Palmer amaranth dispersal and potential establishment into northern United States warrant investigations on species morphology growing in such environments.  Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020].  Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under in corn, soybean and fallow at two timings across five locations in the mid/upper United States Midwest.

### 2021-07-26 Frontiers/Frontiers.tex
as

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
considered one of the most economically damaged weed species to cropping
systems in the United States (Ward et al., 2013). The species has showed
a remarkable capacity to evolve resistance to herbicides. Palmer
amaranth has evolved resistance to eight herbicide sites of action,
increasing the weed management complexity (Lindsay et al., 2017).
Uncontrolled Palmer amaranth in competition for water, light and
nutrients can drastically reduce crop yields (Berger et al., 2015).
Palmer amaranth is documented with potential to reduce 91\%, 68\%, and
X\% of corn (Massinga et al., 2001), soybean (Klingaman and Oliver,
1994), and cotton yields.

Palmer amaranth is a fast growing summer annual forb indigenous to
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with species
biology. The increased use and movement of farm equipment across
locations, conservation agriculture (e.g., no-till), and reliance on
herbicides for weed management are the main human mediated selection of
Palmer amaranth to cropping systems. On the other hand, Palmer amaranth
is a prolific seed producer with a C4 photosynthetic apparatus (Ward et
al., 2013). With a dioecy nature, Palmer amaranth male and female plants
are obligate outcrosser species, increasing the chances of exchanging
herbicide resistant alleles among plants (Oliveira et al., 2018). Also,
Palmer amaranth small seed size (1 mm) tend to thrive in no-tillage
systems (Price et al., 2011), and spread across locations through farm
equipment (Sauer, 1972), manure (Yu et al., 2021), animals (Farmer et
al., 2017), and plant propagules (Hartzler and Anderson, 2016).
Therefore, Palmer amaranth dispersal capacity make the species one of
the most successful cases of weed adaption to cropping systems.

Light and temperature are likely the main environment requirements for
Palmer amaranth successful adaptation. Palmer amaranth is reported with
an extended germination period (Jha et al., 2010). Germination of Palmer
amaranth is triggered by 18 C soil temperature (Keeley et al., 1987),
and optimal germination and biomass production were at 35/30 C day and
night temperatures (Guo and Al-Khatib, 2003). Water has not shown to
limit Palmer amaranth fitness. Under continuous water stress, Palmer
amaranth survived and produced at least 14000 seeds plant-1 (Chahal et
al., 2018). Also, seeds from Palmer amaranth growing under water stress
conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). The continuous global temperature warming can
impact agriculture and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, the greatest
climatic risk of Palmer amaranth establishment is agronomic crops in
Australia and Sub-Sahara Africa (Kistner and Hatfield, 2018).
Temperature is a key factor limiting Palmer amaranth northward range
expansion (Briscoe Runquist et al., 2019); however, under future climate
change Palmer amaranth is likely to expand northward into Canada and
Northern Europe (Kistner and Hatfield, 2018).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the US, Palmer amaranth is commonly found at crop
(Garetson et al., 2019) and non-crop land (Bagavathiannan and
Norsworthy, 2016) in the warm southern United States but its range is
expanding to cool temperatures northward. For example, herbicide
resistant Palmer amaranth is widespread in Nebraska (Oliveira et al.,
2021), Michigan (Kohrt et al., 2017), and Connecticut (Aulakh et al.,
2021). Successful cases of Palmer amaranth invasion and near to
eradication is well documented in Minnesota (Yu et al., 2021) and Iowa.
No Palmer amaranth actively growing was found in Canada; however, Palmer
amaranth seeds was detected in sweet potato slips (Page et al., 2021).
Nonetheless, it seems fated to manage Palmer amaranth in agronomic crops
throughout multiple environments in the near future. Therefore,
strategies on Palmer amaranth management should encompass the
agroecosystem level but not attempts to eradicate the weed. The
continuous Palmer amaranth dispersal and potential establishment into
northern US/Canada warrant investigation on species growing morphology
in such environments.

Most Palmer amaranth studies are based on reactive (e,g, herbicide and
tillage) rather than proactive management. Understanding Palmer amaranth
biology and growing strategies under different environments can enhance
our knowledge on species adaptation. Also, it can aid on designing
proactive and ecological tactics to limit Palmer amaranth range
expansion and reduce its negative impact while conserving diversity.
Therefore, the objective of this study was to investigate the flowering
pattern, biomass production, and height of Palmer amaranth growing under
different environments and timings across five locations.

@@

@@

@@

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
considered one of the most economically damaged weed species to cropping
systems in the United States (Ward et al., 2013). The species has showed
a remarkable capacity to evolve resistance to herbicides. Palmer
amaranth has evolved resistance to eight herbicide sites of action,
increasing the weed management complexity (Lindsay et al., 2017).
Uncontrolled Palmer amaranth in competition for water, light and
nutrients can drastically reduce crop yields (Berger et al., 2015).
Palmer amaranth is documented with potential to reduce 91\%, 68\%, and
54\% of corn (Massinga et al., 2001), soybean (Klingaman and Oliver,
1994), and cotton (Morgan et al., 2001) yields.

Palmer amaranth is a fast growing summer annual forb indigenous to
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with species
biology. Farm mechanization, conservation agriculture (e.g., no-till),
and reliance on herbicides for weed management are the main human
mediated selection of Palmer amaranth to cropping systems. On the other
hand, Palmer amaranth is a prolific seed producer with a C4
photosynthetic apparatus (Ward et al., 2013). With a dioecy nature,
Palmer amaranth male and female plants are obligate outcrosser species,
increasing the chances of exchanging herbicide resistant alleles among
plants (Oliveira et al., 2018). Also, Palmer amaranth small seed size (1
mm) tend to thrive in no-tillage systems (Price et al., 2011), and
spread across locations through farm equipment (Sauer, 1972), manure (Yu
et al., 2021), animals (Farmer et al., 2017), and plant propagules
(Hartzler and Anderson, 2016). Therefore, Palmer amaranth dispersal
capacity make the species one of the most successful cases of weed
adaption to cropping systems.

Light and temperature are likely the main environment requirements for
Palmer amaranth successful adaptation. Palmer amaranth is reported with
an extended germination period (Jha et al., 2010). Germination of Palmer
amaranth is triggered by 18 C soil temperature (Keeley et al., 1987),
and optimal germination and biomass production occur at 35/30 C day and
night temperatures (Guo and Al-Khatib, 2003). Water has not shown to
limit Palmer amaranth fitness. Under continuous water stress, Palmer
amaranth survived and produced at least 14000 seeds plant-1 (Chahal et
al., 2018). Also, seeds from Palmer amaranth growing under water stress
conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). The continuous global temperature warming can
impact agriculture and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, it is estimated
that the greatest climatic risk of Palmer amaranth establishment is
agronomic crops in Australia and Sub-Sahara Africa (Kistner and
Hatfield, 2018). Temperature is a key factor limiting Palmer amaranth
expansion to cooler geographies (Briscoe Runquist et al., 2019);
however, under future climate change Palmer amaranth is likely to expand
northward into Canada and Northern Europe (Kistner and Hatfield, 2018).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the US, Palmer amaranth is commonly found at crop
(Garetson et al., 2019) and non-crop land (Bagavathiannan and
Norsworthy, 2016) in the warm southern United States but its range is
expanding to cool temperatures northward. For example, herbicide
resistant Palmer amaranth is widespread in Nebraska (Oliveira et al.,
2021), Michigan (Kohrt et al., 2017), and Connecticut (Aulakh et al.,
2021). Successful cases of Palmer amaranth invasion and near to
eradication is well documented in Minnesota (Yu et al., 2021) and Iowa.
No Palmer amaranth actively growing was found in Canada; however, Palmer
amaranth seeds was detected in sweet potato slips (Page et al., 2021).
Nonetheless, it seems fated to manage Palmer amaranth in agronomic crops
throughout multiple environments in the near future. Therefore,
strategies on Palmer amaranth management should encompass the
agroecosystem level but not attempts to eradicate the weed. Most tactics
to manage Palmer amaranth are based technology fixes (Scott, 2011),
which are short-term (e.g., herbicide and/or tillage) rather than
long-term Palmer amaranth management.

The continuous Palmer amaranth dispersal and potential establishment
into northern United States warrant investigations on species morphology
growing in such environments. Understanding Palmer amaranth biology and
growing strategies under different agroecossystems can enhance our
knowledge on species adaptation. Also, it can aid on designing proactive
and ecological tactics to limit the species range expansion, reduce its
negative impact, and design resilient and sustainable farming systems
(MacLaren et al., 2020). Therefore, the objective of this study was to
investigate the flowering pattern, biomass production, and height of
Palmer amaranth growing under in corn, soybean and fallow at two timings
across five locations in the mid/upper United States Midwest.

@@

\leavevmode\hypertarget{ref-maclaren2020}{}%
MacLaren, C., Storkey, J., Menegat, A., Metcalfe, H., and
Dehnen-Schmutz, K. (2020). An ecological future for weed science to
sustain crop production and the environment. {A} review. \emph{Agron.
Sustain. Dev.} 40, 24.
doi:\href{https://doi.org/10.1007/s13593-020-00631-6}{10.1007/s13593-020-00631-6}.

@@

\leavevmode\hypertarget{ref-morgan2001}{}%
Morgan, G. D., Baumann, P. A., and Chandler, J. M. (2001). Competitive
{Impact} of {Palmer Amaranth} ({Amaranthus} palmeri) on {Cotton}
({Gossypium} hirsutum) {Development} and {Yield}. \emph{Weed Technology}
15, 408--412.
doi:\href{https://doi.org/10.1614/0890-037X(2001)015\%5B0408:CIOPAA\%5D2.0.CO;2}{10.1614/0890-037X(2001)015{[}0408:CIOPAA{]}2.0.CO;2}.

@@

\leavevmode\hypertarget{ref-scott2011}{}%
Scott, D. (2011). The {Technological Fix Criticisms} and the
{Agricultural Biotechnology Debate}. \emph{J Agric Environ Ethics} 24,
207--226.
doi:\href{https://doi.org/10.1007/s10806-010-9253-7}{10.1007/s10806-010-9253-7}.
